---
title: "Final Project"
author: "Maiko Hata & Michelle Cui"
format: 
  pdf:
    mainfont: "Times New Roman"
    sansfont: "Times New Roman"
fig-width: 7
fig-height: 4
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
bibliography: 
- references.bib
- grateful-refs.bib
---

# Introduction

The Individuals with Disabilities Education Act (IDEA) Part C is a federal grant program that supports states in providing Early Intervention (EI) services for children zero to two who experience developmental delays and/or disabilities (Early Childhood Technical Assistance Center, 2023). These EI services incorporate family-centered practices, aiming to support caregivers' understanding of their children to provide the most effective support for their development (Romano, 2006). However, racially and linguistically minoritized children are less likely than White, English speaking monolingual children to receive EI services (Morgan, et al., 2012) despite federal mandates to ensure the most marginalized, hard-to-reach communities are served.

The IDEA Section 618 requires that data is collected on children with an active IFSP who stopped receiving EI services (U.S. Department of Education \[DOE\], 2024). Recent data revealed that racially and/or linguistically marginalized infants and toddlers are much more likely to leave EI services via disqualification due to non-response to agencies’ outreach efforts after being made eligible.

The purpose of this project is to explore the patterns in which families from racially and/or linguistically marginalized communities leave EI services by examining extant data on EI exit between 2013-2022. The data was obtained from the Office of Special Education Services (OSEP, 2024). The research questions are as follows:

-   Are there associations between children's race and their reasons for exiting EI services nationally, as analyzed through descriptive statistics including chi-square tests and odds ratio analysis?
-   What are the associations between children's race (Black/African American and White) and being disqualified due to "Attempts to Contact Unsuccessful"? What are the odds ratio for being disqualified for this reason between the two racial groups?

# **Methods**

**Independent variables (IV)**: Student' race served as the independent variable (IV), while exit reason served as the dependent variable (DV). There were seven racial categories serving as IVs (Alaska Native/American Indian, Asian, Black/African American, Hispanic, Multiracial, Pacific Islander, White). We briefly discussed collapsing racial categories in order to make the analysis more powerful (as Maiko will also be working on state-level data with cells with "0"s as certain populations are very underrepresented). However, we decided against doing so as this would likely obscure the results, as there are large disparities within BIPOC populations when we look at their exit reasons.

**Dependent variables (DV)**: As you can see in Table 1, there are ten exit categories under three general exit reason “umbrellas” (Hansen et al., 2016):

```{r include=FALSE}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
cite_packages(out.dir = ".")
```

```{r}
exit_categories <- import(here("Data","exit_categories.xlsx")) %>%
 clean_names() 
```

```{r}
kable(exit_categories, 
      caption = "Table of Exit Reasons",
      col.names = c('Exit Reasons','Exit Category Codes')) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

These ten reasons were collapsed into six reasons based on the scope of this study and for logistical reasons. For example, "Deceased" is beyond the scope of this study; one reason is not used in Oregon; multiple codes were similar in nature to each other:

-   Attempts to contact unsuccessful
-   Withdrawal by parent
-   Complete/not eligible for Part B
-   Moved out of state
-   Part B eligibility not determined
-   Part B eligible

**Preparatory work**: We prepared the data in a following manner:

1.  Created an Excel sheet from the national and Oregon data sets
2.  Imported Excel sheet into RStudio
3.  Collapsed/removed DVs
4.  Collapsed data from multiple years into one aggregated data by race

**Data Analysis**: We used chi-square goodness of fit test to understand associations between children's race and their EI exit reasons. Chi-square tests tell us if we can be confident that differences in counts and expected counts are not due to chance. In other words, chi-square tests can be used to evaluate if there is a statistically significant relationship between two dichotomous or nominal variable. However, they are not able to indicate the strength or the direction of the relationship (Morgan et al., 2020).

First, we ran descriptive analysis of the national data set as an omnibus test. For this, we used foundational statistical functions and chi-square to test our null-hypothesis; there is no associations between children's race and their exit reasons.

We then analyzed the association between the exit reason, "Attempts to Contact Unsuccessful", using similar analysis. For this portion, we looked at the association between two racial categories, Black/African American and White infants/toddler groups, with "Attempts to Contact Unsuccessful". We created 2x2 table for this analysis, complete with the total number of exits. This was used to analyze the odd ratio and Cohen's \*h\*. Odds ratio are commonly used for reporting the odds of one outcome between two independent groups (Morgan, et al., 2020).

# Results

```{r include=FALSE}
#|label: load-data
# importing the data and cleaning it up

byrace <- import(here("Data", "race v.2.xlsx")) %>% 
  clean_names() %>% 
  select(-7, -12) %>% 
  mutate(complete_or_not_eligible = complete_prior_to_reaching_max_age_for_part_c + not_eligible_for_part_b_exit_with_no_referrals + not_eligible_for_part_b_exit_with_referrals_to_other_programs)

# str(byrace) I think we can remove this because it is only for examining the structure
```

```{r}
#| label: aggregated-by-race

# aggregated by race and state
agg_by_race_and_state <- byrace %>% 
  group_by(race, area) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent),
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful),
            moved_out_of_state = sum(moved_out_of_state),
            part_b_eligible_exiting_part_c = sum(part_b_eligible_exiting_part_c),
            complete_or_not_eligible = sum(complete_or_not_eligible))

# aggregated by race for OR
race_oregon <- agg_by_race_and_state %>%
  filter(area == "Oregon")

# aggregated by race for US
race_us <- agg_by_race_and_state %>% 
  filter(area == "US and Outlying Areas")
```

```{r}
#| label: chisq-test-trial-omnibus
# run a quick chi square tests for Omnibus. Saving the p-value and test-statistics p value and result 
chi_omni <-chisq.test(race_us[, 3:8])
omni_p_value <- chi_omni$p.value
omni_statistic <- chi_omni$statistic

# chisq.test(race_oregon[, 3:8])
```

```{r}
#| label: exit-total
# Do we need to use this table to report the white and black students percentage? Or we can show the result using an inline code in the Results section

# Created two objective that show percentage and total exits - 1127
# Figured out how to pull percentage, need to check the correctness - 1201
# Added area_by_race, but the computed figures are different from what Maiko put in the results - 1202

area_by_race <- byrace %>% 
  group_by(race) %>% 
  summarize(
    total_exit = sum(exiting_total2),
    .groups = "drop") %>% 
  mutate(percentage = round(total_exit/sum(total_exit),
                            digits = 4)*100)

other_by_race <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(
    total_exit = sum(exiting_total2),
    .groups = "drop"
    ) %>% 
  mutate(percentage = round(total_exit/sum(total_exit), digits = 3)*100)

oregon_by_race <- byrace %>% 
  filter(area == "Oregon") %>% 
  group_by(race) %>% 
  summarize(
    total_exit = sum(exiting_total2),
    .groups = "drop" # .groups = drop: All levels of grouping are dropped.
  ) %>% 
  mutate(percentage = round(total_exit/sum(total_exit), digits = 3)*100)
```

The initial exploration included exit data from 3,310,559 children who exited the EI services between 2013 and 2022 nationally. Approximately `r area_by_race %>% filter(race == "BL") %>% pull(percentage)`% of the children were Black/African American, while `r area_by_race %>% filter(race == "WH") %>% pull(percentage)`% of the children were reported as being White. This shows a possible disproportional representation of children, as census showed that during these years, Black/African American and White children represented approximately 14% for Black/African American children under the age of 18 and between 52% to 49 % for White children nationally (The Annie E. Casey Foundation, 2024).

The chi-square omnibus test indicated that there was a statistically significant association between children's race and their exit reasons, X-squared (30, N = 3,310,559) = `r format(round(omni_statistic, digits = 2), scientific = FALSE)`, with a p-value of `r ifelse(omni_p_value < 0.001, "< 0.001", round(omni_p_value, 3))`

```{r}
# We were curious if PI's (Pacific Islanders) small numbers skewed the chi-square, so we took it out and ran it again. It doesn't seem to have impacted much as P value is still very small with or without PI data. 
race_oregon_subset <- race_oregon %>% 
  filter(race != "PI")
# chisq.test(race_oregon_subset[, 3:8])
```

Looking specifically at the "Attempts to Contact Unsuccessful" category, approximately 13.5% of Black/African American infants and toddlers were disqualified from EI services nationally due to agencies losing contact with families, while only about 5.98% of White children were disqualified for the same reason (Figure 1).

```{r}
# US long and wide  

us_data_long <- race_us %>% 
  pivot_longer(
    cols = 4:8,
    values_to = "values",
    names_to = "names"
  ) %>% 
  group_by(race) %>% 
  mutate(percentage = round(values/exit_total, digits = 4))

us_data_wide <- us_data_long %>% 
  select(-c(exit_total,values)) %>% 
  pivot_wider(
    names_from = names,
    values_from = percentage
  )
```

```{r}
agg_by_area <- byrace %>% 
  group_by(area) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent))

# chisq.test(agg_by_area[, 2:3])

```

```{r}
data_oregon <- byrace %>% 
  filter(area == "Oregon") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent)) %>% 
  mutate(proportion = withdrawal_by_parent / exit_total)

# chi square for oregon 
# chisq.test(data_oregon[, 2:3])
```

```{r}
# data national 

data_national <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent)) %>% 
  mutate(proportion = withdrawal_by_parent / exit_total)

# chisq.test(data_oregon[, 2])
```

```{r}
# figure 1 for DQ data in US   

us_data_wide <- us_data_wide %>% 
  select(race, attempts_to_contact_unsuccessful) %>% 
    mutate(attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful * 100)
```

```{r}
# figure 1 continues 

ggplot(us_data_wide, aes(x = race, y = attempts_to_contact_unsuccessful)) +
  geom_col(stat = "identity", fill = "gray43", position = 'dodge') +
   geom_text(aes(label = attempts_to_contact_unsuccessful),
             position = position_dodge(0.9),
             hjust = -0.05,
             size = 2)+
  labs(
    title = "Figure 1: Unsuccessful Contacts/DQ (U.S.)", 
    x = "Race", 
    y = "Percentage of Exit by Race (%)"
  ) +
  theme_classic()+
  scale_x_discrete(labels = c("Alaska Native/American Indian", "Asian", "Black/African American", "Hispanic", "Multiracial", "Pacific Islander", "White"))+
  coord_flip()

```

```{r}
# remaking a df with exit total and DQ 12/1
us_data_attempts <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful)) %>% 
  mutate(proportion = attempts_to_contact_unsuccessful / exit_total)
```

```{r}
# 12/1 DQ chi-square prep continues SUCCESS!! :) 
us_data_attempts_BLWH <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "BL" | race == "WH") 
```

```{r}
# mutated a column grand_total for all the kiddos exiting for earlier figures 
us_data_attempts <- us_data_attempts %>% 
  mutate(grand_total = sum(exit_total), na.rm = TRUE)
```

```{r}
# now, divide the attempts_to_contact_unsuccessful by grand_total x 100 for % of each race represented in relation to the exit total between 2013-2022. 
us_data_attempts <- us_data_attempts %>% 
  mutate(
    grand_total = sum(exit_total, na.rm = TRUE), 
    portion = (exit_total / grand_total) * 100
  )
```

```{r}
# YAY 12/1 
# chi_dq_BLWH_result <- chisq.test(us_data_attempts_BLWH[, 2:3])
# chi_dq_BLWH_result
```

```{r}
# No need for this chunk anymore as I cleaned it up another way: 
# us_data_wide_DQ_BLWH <- us_data_wide %>% 
#  select(race, attempts_to_contact_unsuccessful) %>% 
#  filter(race == "BL" | race == "WH") %>% 
#  mutate(attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful)

# us_data_wide_DQ_BLWH
```

When we look at the data at state level, the numbers change slightly. Approximately 9.85% of Black/African American infants and toddlers were disqualified from EI services in Oregon due to agencies losing contact with families, while only about 8.03% of White children were disqualified for the same reason in Oregon (Figure 2).

```{r}
# Oregon data long and wide
oregon_data_long <- race_oregon %>% 
  pivot_longer(
  cols = 4:8,
  values_to = "values",
  names_to = "names"
 ) %>% 
  group_by(race) %>% 
  mutate(percentage = round(values/exit_total, digits = 4))

oregon_data_wide <- oregon_data_long %>% 
  select(-c(exit_total,values)) %>% 
  pivot_wider(
    names_from = names,
    values_from = percentage
  )
```

```{r}
# figure 2 for Oregon DQ data 
oregon_data_wide_DQ <- oregon_data_wide %>% 
  select(race, attempts_to_contact_unsuccessful) %>%  
  mutate(
    attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful * 100
  )
```

```{r}
# figure 2 continues 
# rotate x axis text labels - 1202
 ggplot(oregon_data_wide_DQ, 
        aes(x = race, y = attempts_to_contact_unsuccessful)) +
  geom_col(stat = "identity", fill = "gray24", position = 'dodge') +
   geom_text(aes(label = attempts_to_contact_unsuccessful),
             position = position_dodge(0.9),
             hjust = -0.3,
             size = 2.2) +
  labs(
    title = "Figure 2: Unsuccessful Contacts/DQ (Oregon)", 
    x = "Race", 
    y = "Percentage of Exit by Race"
  ) +
  theme_classic() +
  scale_x_discrete(labels = c("Alaska Native/American Indian", "Asian", "Black/African American", "Hispanic", "Multiracial", "Pacific Islander", "White"))+
  coord_flip()
 
 # print(oregon_data_wide_DQ) 
```

```{r}
# Did I just make the summary table of what we decided to look at for this project!?!? Okay I did the same work again but that's okay, GAH 
# WE STILL HAVE TO FILTER ONLY BL AND WHITE BEFORE RUNNING THE CHI SQUARE 
# us_data_DQ <- byrace %>% 
#  filter(area == "US and Outlying Areas") %>% 
#  group_by(race) %>% 
#  summarize(exit_total = sum(exiting_total2), 
#            attempts_to_contact_unsuccessful = 
# sum(attempts_to_contact_unsuccessful))

# Filtered BL and WH - 1202
# Is the chisq_test correct? Need to check this

us_data_DQ <- byrace %>%
 filter(area == "US and Outlying Areas", race %in% c("BL", "WH")) %>%
 group_by(race) %>%
 summarize(exit_total = sum(exiting_total2),
           attempts_to_contact_unsuccessful =
sum(attempts_to_contact_unsuccessful))

chisq_dq <- chisq.test(us_data_DQ[, -1])


```

```{r}
# Used this dataset for making the figure 2
us_data_DQ_proportion <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful)) %>% 
  mutate(proportion = (attempts_to_contact_unsuccessful / exit_total) * 100) 
```

The chi-square indicated that there was a statistically significant association between children being Black/African American or White and them leaving EI due to being disqualified nationally. The chi-square test indicated, X-squared (222556.00, N = 2,088,058), *p* \< 2.2e-16 or 0.0000000000000002 (*p* \< .001).

Because whether or not children were Black/African American or White and whether they were likely to be disqualified from EI services due to "Attempts to Contact Unsuccessful" were both binary variables, we computed an odds ratio as well.

In order to calculate the odds ratio to determine the relative likelihood of the students being disqualified between the two groups, a 2x2 contingency table in a matrix format was created and analyzed. The odds of Black infants and toddlers being disqualified from EI services due to "attempts to contact unsuccessful" were significantly higher than those for White infants and toddlers, with an odds ratio of **2.46** (95% CI \[2.43, 2.48\]). This indicates that Black students were approximately 2.46 times more likely than White students to be disqualified from EI services for this reason.

```{r}
# used wording from Morgan et al. (2020) and ChatGPT 
# prepping the df to mutate a column to have non_dq and delete total 
us_data_wide_DQ_BLWH <- us_data_attempts_BLWH %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)

us_data_wide_DQ_BLWH_nototal <- us_data_wide_DQ_BLWH %>% 
  select(-exit_total)

```

```{r}
#Per chatGPT. The columns and rows switched...? 
us_dq_BLWH_matrix <- matrix( 
  c(56155, 359718, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Black", "White")
  ))

```

```{r}
# ODDS RATIO SUCCESS 
# oddsratio(us_dq_BLWH_matrix)
```

```{r}
#for ODDSRATIO another way  
# library(janitor)
# fisher.test(us_dq_BLWH_matrix)
```

Cohen's *h* was calculated to evaluate the effect size of the analysis**.** The result indicated a small to medium effect size, *h* = 0.259. However, even though effect size shows the magnitude of the difference, it is not necessarily considered to be a direct indication of the importance of the findings (Morgan et al., 2020).

```{r}
# Cohen's h prep 

# black_unsuccessful <- 56155
# black_total <- 359718
# white_unsuccessful <- 99922
# white_total <- 1572263
```


```{r}
# Cohen's h wrong formula 
# p1 <- (black_unsuccessful / black_total)
# p2 <- (white_unsuccessful / white_total)

# h <- 2 * (asin(sqrt(p1)) - asin(sqrt(p2)))
# h
```

```{r}
# Correct cohen's h (-.0.259)
# cohenH(df, observation = "column")
# p1 <- 56155 / (56155 + 359718)
# p1
# p2 <- 99922 / (99922 + 1572263)
# p2
# h <- 2 * (asin(sqrt(p2)) - asin(sqrt(p1)))
# h
```


```{r}
# Cameron helped me make sure

# h <- ES.h(p1, p2)
# h

# pwr.p.test(h = h, n = (black_total + white_total))
```

```{r}
# Another way to get cohen's H, can have NEGATIVE COHEN'S H
# We explored column vs row, but column should be the one we should use, so h = -.259
# library(rcompanion)

# cohenH(us_dq_BLWH_matrix, observation = "column", verbose = TRUE)

# h <- ES.h()
```


```{r}
# CITE ALL THE PACKAGES I USE 
# citation("rcompanion")
```

# Discussion

Our analysis revealed that the odds ratio for Black/African American infants and toddlers to be disqualified from EI services due to "Attempts to Contact Unsuccessful" was 2.46 times higher when compared to their White peers nationally. In addition, state-level data showed smaller disparities between disqiualification rates between Black/African American children and that of their White peers. However, there are many limitations to this descriptive analysis.

First of all, we have to remember that race is not a predictive factor for outcomes. At a quick glance, race seems to be associated with inequity in EI service exit patterns. However, research following the completion of the Human Genome Project has shown that race, from a genetic standpoint, does not contribute to health inequities. Instead, it is the environments experienced by racially minoritized communities that play a significant role (Silverstein, 2015). Silverstein cited Kittles (2015) in order to clarify this: “the bulk of those disparities are not due to any biological difference. The vast majority of health disparities are due to social, behavioral, and environmental components”. Race is merely one of the many descriptors for individuals. 

In addition, as Crenshaw (1989) established in her seminal work, we must take the framework of Intersectionality when conducting a research. This type of oversimplified statistical analysis can contribute to reinforce the status-quo where race is quickly to be blamed, rather than the complex environments and multiple layers of identities that members of marginalized communities live in.

The smaller disparity between racial groups in Oregon in terms of disqualification rate could be due to the state's limited diversity, meaning we simply don't have enough data. This makes it challenging to conduct quantitative studies on marginalized populations, even though research  is so urgently needed for that very reason.

Last but not least, researchers have argued that quantitative methods are inequitable, as "the history of quant methods is inseparable from eugenics movement" (p. 4, Castillo & Strunk, 2024) and that it stems from and reinforces inequity. QuantCrit philosophy are based and expands on the centrality of racism and the lack of neutrality in numbers and categories. Going forward, it would be extremely important to remember these tenets and to approach data collection, categorization and analysis with equity and justice as the central philosophy.

\newpage

```{r}
# We decided to go with just handtyping references as opposed to using only Zotero. Inserting Zotero citaitons in-line and handthyping renders mixed and cannot be fixed to make it alphabetical, and Zotero doesn't get the automatic citations right for many of them, especially non-journal articles and books (e.g. online sources, database, older and scanned articles, slides, guidance docs etc.)

```

# References 

Annie E. Casey Foundation. (2024, July). Child population by race and ethnicity. KIDS COUNT Data Center. <https://datacenter.aecf.org/data/tables/103-child-population-by-race-and-ethnicity#detailed/1/any/false/1095,2048,574,1729,37,871,870,573,869,36/72,66,67,8367,69,70,71,12/423,424> 

Castillo, W. & Strunk, K. (2024, November 15). How to QuantCrit \[PowerPoint slides\]. <https://www.sree.org/critical-perspectives>

Crenshaw, K. (1989). Demarginalizing the intersection of race and sex: A Black feminist critique of antidiscrimination doctrine, feminist theory and antiracist politics. University of Chicago Legal Forum, 1989(1), 139-167. 

Early Childhood Technical Assistance Center \[ecta\], (2023, October 6). *Part C of IDEA*. ecta. <https://ectacenter.org/partc/partc.asp>

Individuals with Disabilities Education Act, 20 U.S.C. § 1400 (2004).

Morgan, G.A., Barrett, K.C., Leech, N.L., & Gloeckner, G.W. (2020). *IBM SPSS for introductory statistics: Use and interpretation.* Routledge.

Morgan, P. L., Farkas, G., Hillemeier, M. M., & Maczuga, S. (2012). Are minority children disproportionately represented in Early Intervention and Early Childhood Special Education? Educational Researcher, 41(9), 339–351. https://doi.org/10.3102/0013189X12459678

Romano, S.D. (2006). Historical perspectives. In G. M. Foley & J.D. Hochman (Eds.), *Mental health in early intervention: Achieving unity in principles and practice* (pp. 33-58). Baltimore: Paul H. Brookes Publishing Company.

Silverstein, J. (2015, April 15). Genes don’t cause racial-health disparities, society does. The Atlantic. <https://www.theatlantic.com/health/archive/2015/04/genes-dont-cause-racial-health-disparities-society-does/389637/> 

\newpage

```{r}
cite_packages(output = "paragraph", out.dir = tempdir())
```

