---
title: "Final Project"
author: "Maiko Hata & Michelle Cui"
format: pdf
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
bibliography: references.bib
---

# Introduction

The Individuals with Disabilities Education Act (IDEA) Part C is a federal grant program that supports states in providing Early Intervention (EI) services for children zero to two who experience developmental delays and/or disabilities (Early Childhood Technical Assistance Center, 2023). These EI services incorporate family-centered practices, aiming to support caregivers' understanding of their children to provide the most effective support for their development (Romano, 2006). However, racially and/or linguistically minoritized children are less likely than White, English speaking monolingual children to receive EI services [@morgan2012].

The IDEA Section 618 requires that the data is collected on children who had an active IFSP who stopped receiving services, or “exited” EI services (U.S. Department of Education \[DOE\], 2024). Recent data revealed that racially and/or linguistically marginalized infants and toddlers are much more likely to leave EI services via disqualification from EI services due to non-response to agencies’ outreach efforts after they had been made eligible. The purpose of this study is to explore the patterns in which families from racially and/or linguistically marginalized communities leave EI services.

# **Methods**

This project examined extant data on EI service exit from 2013-2022, which was obtained online from the Office of Special Education Services (OSEP, 2024). The goal of this study was to understand associations between children's race and exit reasons, which then was compared to the data from Oregon which was also obtained from OSEP (2024).

Variables: Student' race served as the independent variable (IV), while exit reason served as the dependent variable (DV). There were seven IVs (Alaska Native/American Indian, Asian, Black/African American, Hispanic, Multiracial, Pacific Islander, White), and ten DVs that were collapsed into 6 categories:

-   Attempts to contact unsuccessful (DQ)
-   Withdrawal by parent (Withdrawal)
-   Complete/not eligible for Part B (Complete or Not Eligible)
-   Moved out of state (Moved Out)
-   Part B eligibility not determined (Eligibility Not Determined)
-   Part B eligible (Eligible)

Preparatory work: As the original data sets from OSEP contained dependent variables that are beyond the scope of this study (e.g. "deceased"), we prepared the data specified below:

1.  Created an Excel sheet from the national and Oregon datasets
2.  Imported Excel sheet into RStudio
3.  Removed two DVs that are beyond the scope of this study
4.  Combined three similar DVs into one
5.  Collapsed multiple years into one aggregated data

Data Analysis: We used chi-square(X2) goodness of fit test to understand associations between children's race and their EI exit reasons. We used fundamental statistical functions and ran chi-square to test our null-hypothesis, which was that there was no associations between children's races and their exit reasons. For this project, Black/African American and White children and their association with "

# Results

The initial exploration included exit data from 3,310,559 children who exited the EI services between 2013 and 2022 nationally. Approximately \_\_% of the children were Black/African American, while \_\_\_ children reported as being White. The chi-square omnibus test indicated that there was a statistically significant association between children's race and their exit reasons, X-squared (30, N = 3,310,559) = 52218, *p* \< 2.2e-16 or 0.00000000000000022.

Looking specifically at the Attempts to Contact Unsuccessful/DQ category, approximately 13.5% of Black/African American children exited EI per disqualification due to agencies losing contact with families, while only about 5.98% of White children were disqualified from services for the same reason (Figure 1). The chi-square indicated that there was a statistically significant association between children being Black/African American or White and them leaving EI due to being disqualified, (CHI SQUARE VALUES BASED ONLY ON THE us_data_DQ_proportion BUT WITH JUS TBLACK AND WHITE!)

```{r include=FALSE}
library(tidyverse)
library(here)
library(rio)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)

#I think you imported quiet a lot of libraries. Did you use all these packages? I think you can clean out those that were not used (Review from Stephen)

# I carefully went over your all your code blocks.  I think they were thoughtfully executed to achieve the objectives set out in your project (Review from Stephen)

# I think each code block was well executed to achieve the object of the project. Dont have much suggestions. Good work done.
```

```{r include=FALSE}
# importing the data and cleaning it up 
byrace <- import(here("Data", "race v.2.xlsx")) %>% 
  clean_names() %>% 
  select(-7, -12) %>% 
  mutate(complete_or_not_eligible = complete_prior_to_reaching_max_age_for_part_c + not_eligible_for_part_b_exit_with_no_referrals + not_eligible_for_part_b_exit_with_referrals_to_other_programs)

# str(byrace) I think we can remove this because it is only for examining the structure

# This chunk is interesting as you sought to standardize column names, remove columns 7 and 12), and create a new column(complete_or_not_eligible) by aggregating three existing columns (Review from Stephen)
```

```{r}
#| labs: exit-total
# Do we need to use this table to report the white and black students percentage? Or we can show the result using an inline code in the Results section

# Created two objective that show percentage and total exits - 1127

# percentage_by_race <- byrace %>% 
#   filter(area == "US and Outlying Areas") %>% 
#   group_by(race) %>% 
#   summarize(
#     total_exit = sum(exiting_total2),
#     .groups = "drop"
#     ) %>% 
#   mutate(percentage = round(total_exit/sum(total_exit), digits = 3)*100)
# percentage_by_race %>% 
#   gt()

other_by_race <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(
    total_exit = sum(exiting_total2),
    .groups = "drop"
    ) %>% 
  mutate(percentage = round(total_exit/sum(total_exit), digits = 3)*100)

# filtering by area (US and Outlying Areas), grouping by race and  creating percentage column. Code looks good and functional to me (Review from Stephen)

oregon_by_race <- byrace %>% 
  filter(area == "Oregon") %>% 
  group_by(race) %>% 
  summarize(
    total_exit = sum(exiting_total2),
    .groups = "drop" # .groups = drop: All levels of grouping are dropped.
  ) %>% 
  mutate(percentage = round(total_exit/sum(total_exit), digits = 3)*100)

#Rounding off the percentage column could be set to 2 instead of 3 (Review from Stephen)
```

```{r}
# aggregated by race and stte
agg_by_race_and_state <- byrace %>% 
  group_by(race, area) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent),
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful),
            moved_out_of_state = sum(moved_out_of_state),
            part_b_eligible_exiting_part_c = sum(part_b_eligible_exiting_part_c),
            complete_or_not_eligible = sum(complete_or_not_eligible))

# aggregated by race for OR
race_oregon <- agg_by_race_and_state %>%
  filter(area == "Oregon")

# aggregated by race for US
race_us <- agg_by_race_and_state %>% 
  filter(area == "US and Outlying Areas")

#aggregating by state of Oregon and US and Outlyng areas relevant to get an understanding of the data at a closer/specific level of detail (Review from Stephen) 
```

```{r}
# run a quick chi square tests for trial/Omnibus 
chisq.test(race_us[, 3:8])

# chisq.test(race_oregon[, 3:8])
```

```{r}
# We were curious if PI's (Pacific Islanders) small numbers skewed the chi-square, so we took it out and ran it again. It doesn't seem to have impacted much as P value is still very small with or without PI data. 
race_oregon_subset <- race_oregon %>% 
  filter(race != "PI")
# chisq.test(race_oregon_subset[, 3:8])
```

```{r}
# trial bar charts 
# ggplot(oregon_data_long, 
# aes(x = race, y = percentage, fill = race)) +  
#  geom_bar(stat = "identity", position = "dodge") +
#  theme_minimal() +
#  facet_wrap(~ names)
```

```{r}
# US long and wide  

us_data_long <- race_us %>% 
  pivot_longer(
    cols = 4:8,
    values_to = "values",
    names_to = "names"
  ) %>% 
  group_by(race) %>% 
  mutate(percentage = round(values/exit_total, digits = 4))

us_data_wide <- us_data_long %>% 
  select(-c(exit_total,values)) %>% 
  pivot_wider(
    names_from = names,
    values_from = percentage
  )
```

```{r}
agg_by_area <- byrace %>% 
  group_by(area) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent))
# chisq.test(agg_by_area[, 2:3])

```

```{r}
data_oregon <- byrace %>% 
  filter(area == "Oregon") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent)) %>% 
  mutate(proportion = withdrawal_by_parent / exit_total)

# chi square for oregon 
# chisq.test(data_oregon[, 2:3])
```

```{r}
# data national 
data_national <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent)) %>% 
  mutate(proportion = withdrawal_by_parent / exit_total)

# chisq.test(data_oregon[, 2])
```

```{r}
# figure 1 for DQ data in US - HOW CAN I ADD DATA LBELS WITH % ON EACH BAR? 

us_data_wide <- us_data_wide %>% 
  select(race, attempts_to_contact_unsuccessful) %>% 
    mutate(attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful * 100)
```

```{r}
# figure 1 continues 
ggplot(us_data_wide, aes(x = race, y = attempts_to_contact_unsuccessful)) +
  geom_bar(stat = "identity", fill = "darkorange3") +
  labs(
    title = "Figure 1: Unsuccessful Contacts/DQ (U.S.)", 
    x = "Race", 
    y = "Percentage of Exit by Race"
  ) +
  theme_minimal()
```

```{r}
# figure 1 continues 
# I NEED TO KEEP THE TOTAL NUMBER OF EXIT IN THIS CHART TO RUN CHI SQUARE 
us_data_wide_DQ_BLWH <- us_data_wide %>% 
  select(race, attempts_to_contact_unsuccessful) %>% 
  filter(race == "BL" | race == "WH") %>% 
  mutate(attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful)
```

```{r}
# Oregon data long and wide
oregon_data_long <- race_oregon %>% 
  pivot_longer(
  cols = 4:8,
  values_to = "values",
  names_to = "names"
 ) %>% 
  group_by(race) %>% 
  mutate(percentage = round(values/exit_total, digits = 4))

oregon_data_wide <- oregon_data_long %>% 
  select(-c(exit_total,values)) %>% 
  pivot_wider(
    names_from = names,
    values_from = percentage
  )
```

```{r}
# figure 2 for Oregon DQ data 
oregon_data_wide_DQ <- oregon_data_wide %>% 
  select(race, attempts_to_contact_unsuccessful) %>%  
  mutate(
    attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful * 100
  )

```

```{r}
# figure 2 continues 
 ggplot(oregon_data_wide_DQ, 
        aes(x = race, y = attempts_to_contact_unsuccessful)) +
  geom_bar(stat = "identity", fill = "green4") +
  labs(
    title = "Figure 2: Unsuccessful Contacts/DQ (Oregon)", 
    x = "Race", 
    y = "Percentage of Exit by Race"
  ) +
  theme_minimal()
 
 # print(oregon_data_wide_DQ) 
```

```{r}
# do we need a contingency table? 
# c_table <- table(us_data_wide_DQ_BLWH$race, 
# us_data_wide_DQ_BLWH$attempts_to_contact_unsuccessful)
```

```{r}
# Did I just make the summary table of what we decided to look at for this project!?!? 
# WE STILL HAVE TO FILTER ONLY BL AND WHITE BEFORE RUNNING THE CHI SQUARE 
us_data_DQ <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful))
```

```{r}
# and did I just run the chi-square for the dataset!?!? NOPE!
# WE STILL HAVE TO FILTER ONLY BL AND WHITE BEFORE RUNNING THE CHI SQUARE 
chisq.test(us_data_DQ[, 2])
```

```{r}
us_data_DQ_proportion <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful)) %>% 
  mutate(proportion = (attempts_to_contact_unsuccessful / exit_total) * 100) 
```

Review notes: I carefully went over your all your code blocks. I think they were thoughtfully executed to achieve the objectives set out in your project

# Discussion

\newpage

# References

Early Childhood Technical Assistance Center \[ecta\], (2023, October 6). *Part C of IDEA*. ecta. <https://ectacenter.org/partc/partc.asp>

@morgan2012b

Romano, S.D. (2006). Historical perspectives. In G. M. Foley & J.D. Hochman (Eds.), *Mental health in early intervention: Achieving unity in principles and practice* (pp. 33-58). Baltimore: Paul H. Brookes Publishing Company.
